Wind veer in Turbulent Ekman Flow - Part 2: Extrapolation to atmospheric scale separation

### 2024-08-05 write2/manuscript_part2_2024.tex
minor edits

\documentclass[a4paper,11pt]{article}

\usepackage[margin=2.5cm]{geometry}
\usepackage{booktabs} 
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{natbib}
\usepackage{xcolor}
\usepackage{tikz}

\newcommand{\p}{\partial}
\newcommand{\gzm}{{\xi}}%{\gamma z_\text{match}}}
\newcommand{\todo}[1]{\textcolor{red}{$[$#1$]$}}
\newcommand{\RE}{\mathbf{Re}}
\hyphenation{Ol-den-burg}

\DeclareRobustCommand\sampleline[1]{%
  \tikz\draw[#1] (0,0) (0,\the\dimexpr\fontdimen22\textfont2\relax)
  -- (2em,\the\dimexpr\fontdimen22\textfont2\relax);%
}

@@

\author{Hauke Wurps and Cedrick Ansorge}

\date{Draft \today}
\title{Towards a Universal Veering Profile for Turbulent Ekman Flow at arbitrary Reynolds number - Part 2\\
  \normalsize LES and DNS of Turbulent Ekman Flow \\}

@@

\begin{abstract}
  % Introduction % Theoretical profile
  Turbulent Ekman flow is a canonical flow representation of the atmospheric boundary layer, characterized by a logarithmic increase of the wind speed near the surface and a turning of the wind vector aloft. In part I of this work, we derived a formulation of the mean wind velocity profiles of the stationary turbulent Ekman flow based on direct numerical simulation (DNS). Here, we explore the extrapolation of these profiles to atmospheric Reynolds numbers using large-eddy simulation (LES).
  % Case desription and setup
  We compare theoretical profiles of the wind vector for low Reynolds-number turbulent Ekman flow to results from DNS and LES. This necessitates the inclusion of viscous effects, usually neglected in LES, along with modifications to the standard bottom boundary condition.
  % Results and Discussion
  Analysis of the grid-convergence for higher resolution unveils a convergence of LES data towards the theoretical profiles for both intermediate and high Reynolds numbers. Our LES thus confirm that--within the logarithmic layer (the lowest 10\,\% of the boundary layer)--- the turning of the wind vector rotates by approximately one-third of its total rotation across the entirety of the boundary layer.
  % Conclusions
  Such agreement of our theoretical formulation and LES data raises our confidence in the underlying assumptions of the theory and thus reinforces the utility of the theoretical profiles as a reference for intermediate and a quasi-reference for higher Reynolds number simulations.
%

@@

% \todo{start introduction rather with TEF than LES (rather subject than method)}
The interaction of atmospheric boundary layer  (ABL) flow with its lower boundary condition, the surface, is a defining property of the ABL \citep{stull:1988}. Relevant aspects of the interaction with the boundary not only include its heterogeneity \citep{avissar:MWR1989,giorgi:RG1997,claussen:BM1991,garratt:BM1990} and roughness \citep{monin:ARF1970,brutsaert:WRR1975,raupach:AMR1991,kostelecky:JFM2024}, but also a wind veer in favor of the pressure gradient force when the surface is approaced \citep{ekman:AMA1905}. An important tool with capabilities to model the flow-surface interaction is Large-Eddy Simulation (LES). Indeed, LES is widely used to model turbulent flow in the atmospheric boundary layer (ABL) \citep{stoll2020large}. This includes various applications, such as simulating wind turbine and wind farm wakes, and interactions within wind farm clusters \citep{porte2011large,mehta2014large,breton2017survey}. Furthermore, LES is applied in complex environments like mountainous or urban areas \citep{stoll2020large,garcia2018predictive} and to investigate the dispersion of pollution \cite{han2018large}. These applications typically focus on the lower segment of the ABL on the rotating earth, known as the Prandtl or surface layer. In this layer, the vertical wind speed profile commonly exhibits a logarithmic increase, followed by a prominent change of wind direction within the Ekman layer above. Accurate characterization of how wind speed and direction vary with height is of great importance for wind-power forecasting and projection \citep{optis2014moving}.

The simplest representation of the ABL taking into account rotation \emph{and} the quasi non-turbulent, free atmosphere aloft is the turbulent Ekman flow. It is a horizontally homogeneous, statistically stationary boundary-layer flow over a rotating flat surface. Here we consider the problem under neutral stratification, where the potential temperature is constant across the whole domain. Turbulent Ekman flow shows the key characteristics shaping the real ABL: a logarithmic layer and the Ekman spiral. Near the surface, viscous forces dominate the flow. The stationary ABL-height is then defined by the interplay between turbulent growth due to the shear instability of the configuration \citep{lilly1966instability} and rotational suppression of turbulence due to the Coriolis effect. The system's statistical equilibrium is defined by one single parameter, the Reynolds number $\RE$, signifying the scale separation between the largest and smallest scale within the problem. The largest scales are given by shear-induced eddies constrained by the boundary layer's height, while the smallest scales reside within the dissipative range, where strong viscous forces suppress turbulence. The dimensional mean solution of the turbulent Ekman layer is influenced by three parameters: the geostrophic wind $G$, the Coriolis parameter $f$, and the kinematic viscosity $\nu$. These three parameters define the Reynolds number $Re_D = GD/\nu$. With $D=\sqrt{2\nu/f}$ the Ekman layer depth of a laminar flow we write

\begin{align}\label{red}
 Re_D = \sqrt{2}\frac{G}{\sqrt{\nu f}}
\end{align}

The mean solutions of the velocity components are functions of only the Reynolds number and the height above the ground $z$: $u(\RE_D,z)$, $v(\RE_D,z)$ (see e.g. \cite{csanady1967resistance}).

@@

However, it remains uncertain, if the extrapolation of the above framework towards atmospheric $\RE$ is valid. The substantial separation of scales in the ABL renders direct numerical simulation (DNS) of the entire turbulent flow at scale impossible. Within the framework of LES, only the larger eddies are resolved, that contain the main share of kinetic energy, while turbulent mixing below the subfilter-scale is parameterized. This technique allows for the simulation of the ABL at very high (atmospheric) $\RE$.

The accurate representation of mean profiles of velocity, i.e. of first-order statistics of the wind field, is commonly assumed when using LES in a predictive manner  \citep{fedorovich2004convective}. Here, we compare LES outcomes to theoretical solutions to determine the validity of this assumption under different conditions. As LES introduces the grid size as an additional parameter (often understood in relation to a filter scale, cf. \cite{pope2004ten}), we conduct a thorough analysis of its impact on simulation results. \cite{esau2004simulation} and \cite{jiang2018large} have executed similar LES analyses of neutral Ekman layers, however, they introduce an additional parameter by selecting an arbitrary roughness length $z_0$, whereas we derive the parameter $z_0$ corresponding to a fluid with a kinematic viscosity of $\nu = 1.5\cdot10^{-5}$ over flat surface.

Here, we examine three different Reynolds numbers $Re_D$: $1600$, $1.5\cdot10^5$, and $10^6$. While the case $Re_D=1600$ is also investigated by DNS (cf. part 1 of this study), $Re_D=1.5 \cdot 10^5$ and $10^6$ correspond to the scale separation found in typical atmospheric boundary layers. For $Re_D=1600$, we compare DNS and LES results using the exact same forcing and settings. This is the first comparison of this kind for the turbulent Ekman flow. The LES results converge to the DNS results with decreasing grid cell size. This approach raises our confidence in the LES results to a new level for the direct comparison with a resolving simulation. Furthermore, it allows us to extend the insight from DNS to atmospheric scale where an approach through DNS is clearly infeasible for computational constraints. We observe a clear $\RE$ dependency in the LES results, matching the predictions of the theoretical framework.

@@

In this chapter, we introduce the velocity profile for the turbulent Ekman layer developed in part I. 
%Definitions
We align the coordinate Ox with the surface stress, Oz points normal to the surface and normal to Ox, and Oy in the span-wise direction, normal to Oxz.
%The mean velocity profiles for all $\RE$ exhibit a dual scaling nature: in proximity to the lower boundary, the velocity profiles collapse when scaled using
In proximity to the lower boundary, the boundary layer scales in viscous (inner) units, denoted by the index ``$+$'' ($z^+ = zu_\star/\nu$, $U^+=U/u_\star$, where $U$ is the velocity component in x-direction and $u_\star = \sqrt{\nu \p_z U}$ the friction velocity). From the top perspective, the profiles converge when scaled using outer units, denoted by the index ``$-$'' ($z^-=z/\delta$, $U^-=U/G$, where $\delta=u_\star/f$ is the boundary-layer depth scale, $f$ the Coriolis parameter and $G$ the geostrophic wind speed). 
At the lower boundary, the x-axis of the inner units aligns with the shear stress, while the x-axis of the outer units aligns with the geostrophic wind. The angle between both axes is denoted as $\alpha_\star$, the surface veer of the wind across the boundary layer. 
%
Profiles of the velocity components are studied in three different layers reflecting the change of dominant balance when moving away from the surface. These are
(i) the Ekman layer denoted $(\cdot)_\text{Ek}$ (Sec.~\ref{sec:ekmanlayer}), 
(ii) the viscous sub-layer, denoted by an index $(\cdot)_\text{visc}$, 
and (iii) the logarithmic layer denoted by $(\cdot)_\text{log}$
We match $U_\text{visc}$ and $U_\text{log}$ according to their respective formulations and combine them to $U_\text{inner}$ (Sec.~\ref{sec:innerlayer}).
Inner and outer profiles are then matched by a weighting transfer function based on the error function. 
%The span-wise component of the velocity is separated into two layers, namely the inner layer $V_\text{inner}$, and the Ekman layer $V_{EK}$.

@@

\subsection{Ekman layer}
\label{sec:ekmanlayer}
The outer layer of the ABL is characterized by a triadic balance between turbulent flux, pressure gradient and the Coriolis; the vertical change of the Coriolis force causes a pronounced height dependence of the wind direction resulting in the Ekman spiral \citep{ekman:AMA1905}. The classic solution employs Ekman dynamics down to the surface---and thus also in the surface layer, where this is a rather poor representation of turbulent mixing. Here, we model the surface layer by a lower boundary condition for the Ekman spiral determined from DNS:
\begin{subequations}
  \begin{align}
    U_{EK} &= G + Ae^{-\tilde{z}}\cos\tilde{z},\\
		V_{EK} &= - Ae^{-\tilde{z}}\sin\tilde{z},
  \end{align}
\end{subequations}
where the x-axis is aligned with the geostrophic wind and $A = 8.4u_\star$, $\tilde{z} = (z-z_r)/D_E$, $z_r = 0.12\delta$, and $D_E = 3\delta/4\pi\approx 0.24\delta$. The transition from the logarithmic layer to the Ekman layer is located at $z^{-}=0.28- 2.25 Re_{D}^{-1/2}$ with a transition scale of $\sigma_T=2$ for the stream-wise velocity.

@@

\begin{figure}
  \centerline{
	\includegraphics[width=\textwidth]{figures_2024/alt_single_theor_Ekman_profiles_2022.pdf}}
  \caption{a) Theoretical shear-aligned velocity profile ($U^+$,$-V^+$) of the turbulent Ekman flow for $Re_D = 2000$ (\sampleline{}) and $Re_D = 4000$ (\sampleline{dashed}) and the different layers (colors). b) Hodograph of the geostrophy-aligned velocity components}
  \label{single_ekman}. 
\end{figure}

@@

\section{Case description and numerical set-up}
\label{setup}
%
%\subsection{Notation}
%
%\begin{table*}
	%\centering
%%	\caption{}
	%\begin{tabular}{ll}
    %\toprule
		%$\Delta$ & grid cell size in LES \\
		%$\overline{u''w''}_0$ &  \\
		%$\overline{v''w''}_0$ &  \\
		%$\delta_{95}$ & height where the turbulent stress decreases to 5\% of its surface value\\
		%\bottomrule
	%\end{tabular}
%\end{table*}

@@

\begin{table*}
	\centering
	\caption{Parameters of the simulated cases. $f$, $\nu$, and $G$ are input parameters and define the Reynolds numbers $Re_D$ and $Re_\tau$, while $u_\star$, $\alpha_\star$, and $\delta = u_\star/f$ are resulting properties of the flow}
	\begin{tabular}{lcccccccc}
          \toprule  
	  Name & $Re_D$ & $Re_\tau$ & $f$ [s$^{-1}$] & $\nu$ [m$^2$s$^{-1}$] & $G$ [ms$^{-1}$] & $u_\star$ [ms$^{-1}$] & $\alpha_\star$ [$^\circ$]& $\delta$ [m] \\
          \midrule
	  Re1 & $1.6\times10^3$ & $3.0\times10^3$ &  &  & 0.0438 & 0.00211 & 16.8 & 21.1 \\
	  Re2 & $1.5\times10^5$ & $7.3\times10^6$ & $10^{-4}$ & $1.5\times10^{-5}$ & 4.108 & 0.1048 & 8.5 & 1048\\
	  Re3 & $1\times10^6$ & $2.2\times10^8$ &  &  & 27.39 & 0.5785 & 7.0 & 5785 \\
          \bottomrule
	\end{tabular}
	\label{simulation_parameters}
\end{table*}
An incompressible, turbulent Ekman flow over a hydrodynamically smooth surface is simulated using the Parallel Large Eddy Simulation model (PALM, \citealp{maronga2020overview}). We study three different Reynolds numbers Re$_D = 10^3;1.5\times 10^5;10^6$, hereafter Re1, Re2, and Re3, respectively (Tab.~\ref{simulation_parameters}).
%
The domain is rotating around the z-axis with an angular velocity corresponding to the Coriolis parameter $f=10^{-4}s^{-1}$. The stratification of the flow is truly neutral, i.e., the potential temperature is constant across the whole domain. A constant boundary layer height on the order of $\delta=u_\star/f$ forms due to the balance between shear production and rotational suppression of turbulence. At the upper boundary, the flow is non-turbulent. Hence, a no-penetration boundary condition is used for the vertical velocity component while the horizontal components are prescribed according to geostrophic balance--a Dirichlet-type boundary condition. At the bottom, a constant-flux layer is assumed and Monin--Obukhov similarity theory (MOST) is used to calculate the surface momentum fluxes. The Navier-Stokes equations are integrated using a 3$^\text{rd}$-order low-storage Runge--Kutta method. For scalar advection a 5$^\text{th}$-order Wicker--Skamarock scheme is employed. The Poisson equation is solved using a direct fast Fourier transform (FFT). In LES, the turbulent transport on the subgrid scale (SGS) needs to be modeled by an SGS closure model. We use two kinds of SGS closures to assess their impact on the LES solution: a 1.5-order closure after \cite{deardorff1980stratocumulus} and a dynamic closure after \cite{heinz2008realizability}. For most of the simulations, the 1.5-order closure is used, since the dynamic closure needs more computational resources, but the simulations with $\Delta^-=200^{-1}$ are repeated using the dynamic closure.

@@

In part I of this publication, the DNS of $Re_D\leq1\,600$ is carried out for a horizontal domain size of $(0.54\Lambda_{Ro})^2$ - $(1.08\Lambda_{Ro})^2$, where $\Lambda_{Ro}=G/f$ is the Rossby radius. For these Reynolds numbers, $Z\approx 0.05$, such that $L_x \approx 10 \delta$. However, $Z$ decreases logarithmically with increasing Reynolds number and it is $Z\approx 0.02$ for $Re_D=10^6$. A domain size of half the Rossby radius would then extend to $L_x\approx 25\delta$. Such a large domain would imply immense computational cost. \cite{spalart2008direct} used a horizontal domain of $L_x = 2\delta$, arguing that this length allows the resolution of the largest outer-layer eddies according to \cite{csanady1967resistance}. During a sensitivity test of the domain size we observed that simulations with domain sizes $L\geq 4\delta$ often tend to accumulate turbulence kinetic energy in the upper half of the the boundary layer. This TKE increases over several inertial oscillations with energy mostly on the scale of the domain size. Such a development was not observed in the DNS. We could successfully avoid such a behaviour by using a domain of size $L \approx 3\delta$ in combination with a shifted periodic boundary condition in y-direction, as described by \cite{munters2016shifted}. Although for the Ekman flow the direction of the mean flow is only aligned with the x-direction of the grid near the surface, a shift of the boundary condition by $L/3$ significantly helped to suppress the accumulation of TKE in the upper half of the boundary layer.

@@

In LES at low $\RE$ or very high resolution the subgrid eddy viscosities may fall far below the molecular viscosity of air. In the context of PALM's Deardorff closure, it is $K_m = c_0 \Delta \sqrt{e}$, where $c_0 = 0.1$ \citep{deardorff1980stratocumulus}, $\Delta$ is the grid size and $e$ is the SGS-TKE, calculated by a prognostic equation. Hence, very low $e$ as well as fine resolution can lead to $K_m<\nu$. When this is the case, $\nu$ cannot be ignored anymore, and we let $K_m = c_0 \Delta \sqrt{e} + \nu$. Hence, (absent stratification) the governing equation of for momentum in our version of PALM is
\begin{align}\label{gov_palm}
  \frac{\partial u_i}{\partial t} = - \frac{\partial u_i u_j}{\partial x_j} -\epsilon_{ijk}f_ju_k + \epsilon_{i3j} f_3 u_{g,j} - \frac{1}{\rho_0}\frac{\partial\pi^*}{\partial x_i} + \frac{\partial}{\partial x_j}\left( K_m\left(\frac{\partial u_i}{\partial x_j} + \frac{\partial u_j}{\partial x_i} \right)\right).
\end{align}
In the limit of either well-resolved simulations or low $\RE$ (or both), $K_m \rightarrow \nu$, so the last term on the right hand side of eq. (\ref{gov_palm}) becomes $\nu\frac{\partial^2u_i}{\partial x_j^2}$, which complies with the Navier-Stokes equations for an incompressible fluid underlying our reference solution from DNS.

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width=0.7\textwidth]{figures_2024/d3y_3Re_hodograph}
  \caption{Geostrophy aligned hodographs of the LES in comparison to the theoretical and laminar hodographs. A $\times$ indicates the first grid point of the LES. At the points marked with $+$, the velocity is used to determine $u_\star$}
  \label{3Re_hodograph}
\end{figure}

>>>>

\documentclass[smallcondensed,draft]{svjour3}

\usepackage{lmodern}
%\usepackage[margin=2.5cm]{geometry}
\usepackage{booktabs} 
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
%\usepackage{caption}
\usepackage[authoryear]{natbib}
\usepackage{xcolor}
\usepackage{tikz}

\usepackage{adjustbox}

\newcommand{\p}{\partial}
\newcommand{\gzm}{{\xi}}%{\gamma z_\text{match}}}
\newcommand{\todo}[1]{\textcolor{red}{$[$#1$]$}}
\newcommand{\RE}{\mathbf{Re}}

\hyphenation{Ol-den-burg}
\hyphenation{non-sta-tio-na-ri-ty}
\DeclareRobustCommand\sampleline[1]{%
  \tikz\draw[#1] (0,0) (0,\the\dimexpr\fontdimen22\textfont2\relax)
  -- (2em,\the\dimexpr\fontdimen22\textfont2\relax);%
}

@@

\author{Hauke Wurps $^{1}$ and Cedrick Ansorge $^{2}$}
\authorrunning{H. Wurps and C. Ansorge}

\institute{$^{1}$ Uni Oldenburg \newline 
$^{2}$ FU Berlin, Institut f\"ur Meteorologie, Carl-Heinrich-Becker-Weg 6--10, 12165 Berlin \email{cedrick@posteo.de}}

\date{Submitted \today}
\title{Wind veer in Turbulent Ekman Flow - Part 2: Extrapolation to atmospheric scale separation}
\titlerunning{A Universal veering profile for Ekman flow - Part 2}

@@

\begin{abstract}
  % Introduction % Theoretical profile
  The atmospheric boundary layer is characterized by a logarithmic increase of the wind speed in the surface layer and a turning of the wind vector aloft. The simplest canonical configuration featuring both \emph{a logarithmic layer and the wind turning} is Ekman flow. A formulation of the mean velocity profile was derived from scaling considerations and calibrated using direct numerical simulation in part I of this work. Here, we explore the extrapolation to atmospheric Reynolds number (Re) using large-eddy simulation (LES).
  % Case desription and setup
  Theoretical profiles of the wind vector are compared to the theoretical reference at intermediate Re, which necessitates consideration of viscous effects % [not required in abstract] (usually neglected in LES) 
  along with modifications to the standard bottom boundary condition.
  % Results and Discussion
  Analysis of the grid convergence for higher resolution unveils a convergence of LES data towards the theoretical profiles for intermediate and high Re. The LES thus confirm that (i) the spanwise velocity scales as $\left.Re_\tau\right.^{-1}$, where $Re_\tau$ is the friction Reynolds number, and (ii) one third of the wind veer is confined to the surface layer.
  % [this level of detail is not needed in the abstract] (the lowest 10\,\% of the boundary layer) 
  % Conclusions
  Such agreement of the theoretical formulation and LES data raises confidence in the underlying scaling assumptions and thus reinforces the utility of the theoretical profiles as a reference for intermediate and a quasi-reference for both idealized simulation and field observation.
%

@@

The interaction of atmospheric boundary layer  (ABL) flow with its lower boundary condition, the surface, is a defining property of the ABL \citep{stull:1988}. Relevant aspects of the interaction with the boundary not only include its heterogeneity \citep{avissar:MWR1989,giorgi:RG1997,claussen:BM1991,garratt:BM1990} and roughness \citep{monin:ARF1970,brutsaert:WRR1975,raupach:AMR1991,kostelecky:JFM2024}, but also a wind veer in favor of the pressure gradient force when the surface is approached \citep{ekman:AMA1905}.

Large-Eddy Simulation (LES) is widely used to study the turbulent flow in the atmospheric boundary layer (ABL) for a wide range of applications \citep{stoll2020large}. This includes the simulation of wind turbines and wind farm wakes, their interactions within wind farm clusters \citep{porte2011large,mehta2014large,breton2017survey}. Furthermore, LES is applied in complex environments like mountainous or urban areas \citep{stoll2020large,garcia2018predictive} and to investigate the dispersion of pollution \citep{han2018large}. These applications typically focus on the lower segment of the ABL on the rotating earth, known as the Prandtl or surface layer. In this layer, the vertical wind speed profile commonly exhibits a logarithmic increase, followed by a prominent change of wind direction within the Ekman layer above. Accurate characterization of how wind speed and direction vary with height is of great importance for wind-power forecasting and projection \citep{optis2014moving}.

The simplest representation of the ABL taking into account rotation and the quasi non-turbulent, free atmosphere aloft is Ekman flow. It is a horizontally homogeneous, statistically stationary boundary-layer flow over a rotating flat surface. Here we consider the problem under neutral stratification, where density effects on the flow due to temperature fluctuations are absent. This turbulent flow configuration features the key characteristics of the real ABL, a logarithmic layer and the Ekman spiral; in immediate vicinity of the surface, viscous forces dominate (\emph{viscous sublayer}). The stationary ABL-height is set by the interplay between turbulent growth due to the shear instability \citep{lilly1966instability} and rotational suppression of turbulence due to the Coriolis effect. The system's statistical equilibrium is characterized by a single parameter, the Reynolds number $\RE$, signifying the scale separation between the largest and smallest scales of the problem. The largest scales are given by shear-induced eddies constrained by the boundary-layer height, while the smallest scales reside within the dissipative range, where strong viscous forces suppress turbulence. The dimensional mean solution of the turbulent Ekman layer is influenced by three parameters: the geostrophic wind $G$, the Coriolis parameter $f$, and the kinematic viscosity $\nu$. These three parameters define the Reynolds number $Re_D = GD/\nu$. Using $D=\sqrt{2\nu/f}$, the Ekman layer depth of a laminar flow, we write
%
\begin{align}\label{red}
 Re_D = \sqrt{2}\frac{G}{\sqrt{\nu f}}
\end{align}
%
The mean solutions of the velocity components are functions of only the Reynolds number and the height above the ground $z$: $u(\RE_D,z)$, $v(\RE_D,z)$ (see e.g. \cite{csanady1967resistance}).

@@

However, it remains uncertain, if the extrapolation of the above framework towards atmospheric $\RE$ is valid. The substantial separation of scales in the ABL makes direct numerical simulation (DNS) of the entire turbulent flow at geophysical scale impossible. Within the framework of LES, only the larger eddies are resolved, that contain the main share of kinetic energy, while turbulent mixing below the subfilter-scale is parameterized. This technique allows for the simulation of the ABL at very high (atmospheric) $\RE$.

The accurate representation of mean profiles of velocity, i.e. of first-order statistics of the wind field, is commonly assumed when using LES in a predictive manner  \citep{fedorovich2004convective}. Here, we compare LES outcomes to theoretical solutions to determine the validity of this assumption under different conditions. As LES introduces the grid size as an additional parameter (often understood in relation to a filter scale, cf. \cite{pope2004ten}), we analyse its impact on simulation results. \cite{esau2004simulation} and \cite{jiang2018large} have executed similar LES analyses of neutral Ekman layers, however, they introduce an additional parameter by selecting an arbitrary roughness length $z_0$, whereas we derive the parameter $z_0$ corresponding to a fluid with a kinematic viscosity of $\nu = 1.5\cdot10^{-5}$ over a flat surface.

Here, we examine three different Reynolds numbers $Re_D$: $1600$, $1.5\cdot10^5$, and $10^6$. While the case $Re_D=1600$ is also investigated by DNS (cf. part 1 of this study), $Re_D=1.5 \cdot 10^5$ and $10^6$ correspond to the scale separation found in typical atmospheric boundary layers. For $Re_D=1600$, we compare DNS \citep{ansorge:2024b} and LES results using the exact same forcing and settings. This is the first comparison of this kind for the turbulent Ekman flow.

@@

In this chapter, we introduce the velocity profile for the turbulent Ekman layer developed in part I. 
%Definitions
We align the coordinate Ox with the surface stress (\cite{ansorge:BM2019}), Oz points normal to the surface and normal to Ox, and Oy in the span-wise direction, normal to Oxz. In proximity to the lower boundary, the boundary layer scales in viscous (inner) units, denoted by the index ``$+$'' ($z^+ = zu_\star/\nu$, $U^+=U/u_\star$, where $U$ is the velocity component in x-direction and $u_\star = \sqrt{\nu \p_z U}$ the friction velocity). From the top perspective, the profiles converge when scaled using outer units, denoted by the index ``$-$'' ($z^-=z/\delta$, $U^-=U/G$, where $\delta=u_\star/f$ is the boundary-layer depth scale, $f$ the Coriolis parameter and $G$ the geostrophic wind speed). 
At the lower boundary, the x-axis of the inner units aligns with the negative shear stress, while the x-axis of the outer units aligns with the geostrophic wind. The angle between both axes is denoted as $\alpha_\star$, the surface veer of the wind across the boundary layer. 
%
Profiles of the velocity components are studied in three different layers reflecting the change of dominant balance when moving away from the surface. These are
(i) the Ekman layer denoted $(\cdot)_\text{Ek}$ (Sec.~\ref{sec:ekmanlayer}), 
(ii) the viscous sub-layer, denoted by an index $(\cdot)_\text{visc}$, 
and (iii) the logarithmic layer denoted by $(\cdot)_\text{log}$
We match $U_\text{visc}$ and $U_\text{log}$ according to their respective formulations and combine them to $U_\text{inner}$ (Sec.~\ref{sec:innerlayer}).
Inner and outer profiles are then matched by a weighting transfer function based on the error function. 
%The span-wise component of the velocity is separated into two layers, namely the inner layer $V_\text{inner}$, and the Ekman layer $V_{EK}$.

@@

\subsection{Ekman layer}
\label{sec:ekmanlayer}
The outer layer of the ABL is characterized by a triadic balance between turbulent flux, pressure gradient and the Coriolis force; the vertical change of the Coriolis force causes a pronounced height dependence of the wind direction resulting in the Ekman spiral \citep{ekman:AMA1905}. The classic solution employs Ekman dynamics down to the surface---and thus also in the surface layer, where this is a rather poor representation of turbulent mixing. Here, we model the surface layer by a shifted lower boundary condition for the Ekman spiral determined from DNS:
\begin{subequations}
  \begin{align}
    U_{EK} &= G + Ae^{-\tilde{z}}\cos\tilde{z},\\
		V_{EK} &= - Ae^{-\tilde{z}}\sin\tilde{z},
  \end{align}
\end{subequations}
where the x-axis is aligned with the geostrophic wind and $A = 8.4u_\star$, $\tilde{z} = (z-z_r)/D_E$, $z_r = 0.12\delta$, and $D_E = 3\delta/4\pi\approx 0.24\delta$. The transition from the logarithmic layer to the Ekman layer is located at $z^{-}=0.28- 2.25 Re_{D}^{-1/2}$ with a transition scale of $\sigma_T=2$ for the stream-wise velocity.

@@

\begin{figure}
  \centerline{
	\includegraphics[width=\textwidth]{figures/alt_single_theor_Ekman_profiles_2024.png}}
  \caption{a) Theoretical shear-aligned velocity profile ($U^+$,$-V^+$) of the turbulent Ekman flow for $Re_D = 2000$ (\sampleline{}) and $Re_D = 4000$ (\sampleline{dashed}) and the different layers (colors). b) Hodograph of the geostrophy-aligned velocity components}
  \label{single_ekman}. 
\end{figure}

@@

\section{Case description and numerical set-up}
\label{setup}

@@

\begin{table*}
	\centering
	\caption{Parameters of the simulated cases. $f$, $\nu$, and $G$ are input parameters and define the Reynolds numbers $Re_D$ and $Re_\tau$, while $u_\star$, $\alpha_\star$, and $\delta = u_\star/f$ are resulting properties of the flow}
	\begin{adjustbox}{width=\textwidth}
    \begin{tabular}{l|ccccc|ccc}
          \toprule
	  Name & $Re_D$ & $Re_\tau$ & $f$ [s$^{-1}$] & $\nu$ [m$^2$s$^{-1}$] & $G$ [ms$^{-1}$] & $u_\star$ [ms$^{-1}$] & $\alpha_\star$ [$^\circ$]& $\delta$ [m] \\
          \midrule
	  Re1 & $1.6\times10^3$ & $3.0\times10^3$ &  &  & 0.0438 & 0.00211 & 16.8 & 21.1 \\
	  Re2 & $1.5\times10^5$ & $7.3\times10^6$ & $10^{-4}$ & $1.5\times10^{-5}$ & 4.108 & 0.1048 & 8.5 & 1048\\
	  Re3 & $1\times10^6$ & $2.2\times10^8$ &  &  & 27.39 & 0.5785 & 7.0 & 5785 \\
          \bottomrule
	  \end{tabular}
  \end{adjustbox}
	\label{simulation_parameters}
\end{table*}
An incompressible, turbulent Ekman flow over a hydrodynamically smooth surface is simulated using the Parallel Large Eddy Simulation model (PALM, \citealp{maronga2020overview}). We study three different Reynolds numbers Re$_D = 1600;1.5\times 10^5;10^6$, hereafter Re1, Re2, and Re3, respectively (Tab.~\ref{simulation_parameters}).
%
The domain is rotating around the z-axis with an angular velocity corresponding to the Coriolis parameter $f=10^{-4}s^{-1}$ (the y-component of the Coriolis acceleration was set to zero). The stratification of the flow is truly neutral, i.e., the potential temperature is constant across the whole domain. A constant boundary layer height on the order of $\delta=u_\star/f$ forms due to the balance between shear production and rotational suppression of turbulence. At the upper boundary, the flow is non-turbulent. Hence, a no-penetration boundary condition is used for the vertical velocity component while the horizontal components are prescribed according to geostrophic balance--a Dirichlet-type boundary condition. Below the first grid point, a constant-flux layer is assumed and Monin--Obukhov similarity theory (MOST) is used to calculate the surface momentum fluxes. The Navier-Stokes equations are integrated using a 3$^\text{rd}$-order low-storage Runge--Kutta method. For scalar advection a 5$^\text{th}$-order Wicker--Skamarock scheme is employed. The Poisson equation is solved using a direct fast Fourier transform (FFT). In LES, the turbulent transport on the subgrid scale (SGS) needs to be modeled by an SGS closure model. We use two kinds of SGS closures to assess their impact on the LES solution: a 1.5-order closure after \cite{deardorff1980stratocumulus} and a dynamic closure after \cite{heinz2008realizability}. For most of the simulations, the 1.5-order closure is used, since the dynamic closure needs more computational resources, but the simulations with $\Delta^-=200^{-1}$ are repeated using the dynamic closure.

@@

In part I of this publication, the DNS of $Re_D\leq1\,600$ is carried out for a horizontal domain size of $(0.54\Lambda_{Ro})^2$ - $(1.08\Lambda_{Ro})^2$, where $\Lambda_{Ro}=G/f$ is the Rossby radius. For these Reynolds numbers, $Z\approx 0.05$, such that $L_x \approx 10 \delta$. However, $Z$ decreases logarithmically with increasing Reynolds number and it is $Z\approx 0.02$ for $Re_D=10^6$. A domain size of half the Rossby radius would then extend to $L_x\approx 25\delta$. Such a large domain would imply immense computational cost. \cite{spalart2008direct} used a horizontal domain of $L_x = 2\delta$, arguing that this length allows the resolution of the largest outer-layer eddies according to \cite{csanady1967resistance}. During a sensitivity test of the domain size we observed that simulations with domain sizes $L\geq 4\delta$ often tend to accumulate turbulence kinetic energy in the upper half of the boundary layer. The TKE increases over several inertial oscillations with energy mostly on the scale of the domain size. Such a development was not observed in the DNS. We could successfully avoid such a behavior by using a domain of size $L \approx 3\delta$ in combination with a shifted periodic boundary condition in y-direction, as described by \cite{munters2016shifted}. Although for the Ekman flow the direction of the mean flow is only aligned with the x-direction of the grid near the surface, a shift of the boundary condition by $L/3$ significantly helped to suppress the accumulation of TKE in the upper half of the boundary layer.

@@

In LES at low $\RE$ or very high resolution the subgrid eddy viscosity may fall far below the molecular viscosity of air. In the context of PALM's Deardorff closure, it is $K_m = c_0 \Delta \sqrt{e}$, where $c_0 = 0.1$ \citep{deardorff1980stratocumulus}, $\Delta$ is the grid size and $e$ is the SGS-TKE, calculated by a prognostic equation. Hence, very low $e$ as well as fine resolution can lead to $K_m<\nu$. When this is the case, $\nu$ cannot be ignored anymore, and we let $K_m = c_0 \Delta \sqrt{e} + \nu$. Hence, assuming absent stratification, the governing equation of for momentum in our version of PALM is
\begin{align}\label{gov_palm}
  \frac{\partial u_i}{\partial t} = - \frac{\partial u_i u_j}{\partial x_j} -\epsilon_{ijk}f_ju_k + \epsilon_{i3j} f_3 u_{g,j} - \frac{1}{\rho_0}\frac{\partial\pi^*}{\partial x_i} + \frac{\partial}{\partial x_j}\left( K_m\left(\frac{\partial u_i}{\partial x_j} + \frac{\partial u_j}{\partial x_i} \right)\right),
\end{align}
where $u_i$ is the filtered velocity, $\epsilon_{ijk}$ the Levi-Civita symbol, $f_i$ the Coriolis acceleration, $u_{g,i}$ the geostrophic wind, $\rho_0$ the density of dry air, and $\pi^*$ the modified perturbation pressure. In the limit of either well-resolved simulations or low $\RE$ (or both), $K_m \rightarrow \nu$, so the last term on the right hand side of eq. (\ref{gov_palm}) becomes $\nu\frac{\partial^2u_i}{\partial x_j^2}$, which complies with the Navier-Stokes equations for an incompressible fluid underlying our reference solution from DNS.

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width=0.7\textwidth]{figures/d3y_3Re_hodograph}
  \caption{Geostrophy aligned hodographs of the LES in comparison to the theoretical and laminar hodographs. A $\times$ indicates the first grid point of the LES. At the points marked with $+$, the velocity is used to determine $u_\star$}
  \label{3Re_hodograph}
\end{figure}
